PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-04-20 peace/papers/NAR_web_server_2009/paper.tex
Modifications to paper.

\begin{abstract}
  \mc{We present \peace, a stand alone tool for high-throughput {\it
      ab initio} clustering of transcript fragment sequences produced
    by Next Generation or Sanger Sequencing technologies.  It is
    freely available from
    \href{http://www.peace-tools.org}{www.peace-tools.org}.  Installed
    and managed through a download-able user-friendly GUI, \peace\/
    can process large data sets of transcript fragments of length 50
    bases or greater, grouping the fragments by gene associations with
    a sensitivity compatible to leading clustering tools.  Once
    clustered, the user can employ the GUI's analysis functions,
    facilitating easy collection of statistics and single out specific
    clusters for more comprehensive study or assembly.  Using a novel
    minimum spanning tree based clustering method, \peace is the equal
    of other leading tools in the literature.  It produces results of
    quality virtually identical to those of the \wcd\/ tool when
    applied to Sanger sequences, significantly improved results over
    \wcd\/ when applied the products of Next Generation Sequencing
    Technology, and significantly improved results over \capthree\/ in
    both cases.  In short, \peace\/ provides an intuitive graphical
    user interface and a feature-rich, parallel clustering engine that
    collectively are as good, if not better, than the leading
    clustering tools in the literature.}

@@

\begin{figure}
  \centerline{\includegraphics[width=3in]{pics.d/workflow.png}}
  \caption{Overview of the procedure for clustering and analysis using
    PEACE.}\label{fig:workflow}
\end{figure}

@@

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.pdf}}
  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
    Type 1 error, based on the average over 30 simulated Sanger
    Sequence ESTs sets derived from 100 zebra fish genes (see
    Supplementary Materials, Section C, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd \mc{(version 0.5.1)}, Black/Dot-Dash =
    \capthree; vertical tics = 95\% confidence intervals on estimates.
    Intervals are not presented for Type 1 error due to the effective
    lack of variance.}\label{SeJiT1T2}
\end{figure}

\begin{figure}
   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.pdf}}
   \caption{Comparisons of Runtime: On left, we compare the sequential
     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
     runtime is not reported, as the time spent on clustering cannot
     be differentiated from the time spent on assembly.)  On right,
     a comparison when run in parallel, ranging the number of
     processors from 2 to 12 while holding the EST/Processor ratio
     steady at the constant 8888.  All values represent the average
     of 30 runs; vertical tics = 95\% confidence intervals on
     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
     cache and 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
\end{figure}

\mc{In applying the tools to real data Sanger data, we used the Human
  Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
  the A076941.fa {\it Arabidopsis thaliana} data set used to test WCD
  \cite{Hazelhurst08a,Hazelhurst08b}, (see Table~\ref{SangerReal}).
  We notice essentially identical results for PEACE and WCD in
  quality, both significantly better than Cap3 in Sensitivity and Type
  1 error rate, while slightly worse in the Jaccard Index and Type 2
  error.  In runtime we see some inconsistency, with PEACE showing a
  60\% runtime improvement in the first data set, but requiring 20\%
  more time in the Arabidopsis data set.  In Section~D.1 of the
  Supplementary Materials we present runtimes several more sets,
  observing that while \peace\/ appears to be significantly more
  faster on the smaller sets, \wcd\/ does overtake it for larger
  sets.}

>>>>

\begin{abstract}
  \mc{We present \peace, a stand alone tool for high-throughput {\it
      ab initio} clustering of transcript fragment sequences produced
    by Next Generation or Sanger Sequencing technologies.  It is
    freely available from
    \href{http://www.peace-tools.org}{www.peace-tools.org}.  Installed
    and managed through a download-able user-friendly GUI, \peace\/
    can process large data sets of transcript fragments of length 50
    bases or greater, grouping the fragments by gene associations with
    a sensitivity compatible to leading clustering tools.  Once
    clustered, the user can employ the GUI's analysis functions,
    facilitating easy collection of statistics and allowing them to
    single out specific clusters for more comprehensive study or
    assembly.  Using a novel minimum spanning tree based clustering
    method, \peace\/ is the equal of other leading tools in the
    literature in quality and interface.  It produces results of
    quality virtually identical to those of the \wcd\/ tool when
    applied to Sanger sequences, significantly improved results over
    \wcd\/ when applied the products of Next Generation Sequencing
    Technology, and significantly improved results over \capthree\/ in
    both cases.  In short, \peace\/ provides an intuitive graphical
    user interface and a feature-rich, parallel clustering engine that
    proves to be a valuable addition to the leading cDNA clustering
    tools.}

@@

\begin{figure}
  \centerline{\includegraphics[width=3in]{pics.d/workflow.png}}
  \caption{Overview of the procedure for clustering and analysis using
    \peace.}\label{fig:workflow}
\end{figure}

@@

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.png}}
  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
    Type 1 error, based on the average over 30 simulated Sanger
    Sequence ESTs sets derived from 100 zebra fish genes (see
    Supplementary Materials, Section C, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd\/ \mc{(version 0.5.1)}, Black/Dot-Dash =
    \capthree; vertical tics = 95\% confidence intervals on estimates.
    Intervals are not presented for Type 1 error due to the effective
    lack of variance.}\label{SeJiT1T2}
\end{figure}

\begin{figure}
   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.png}}
   \caption{Comparisons of Runtime: On left, we compare the sequential
     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
     runtime is not reported, as the time spent on clustering cannot
     be differentiated from the time spent on assembly.)  On right,
     a comparison when run in parallel, ranging the number of
     processors from 2 to 12 while holding the EST/Processor ratio
     steady at the constant 8888.  All values represent the average
     of 30 runs; vertical tics = 95\% confidence intervals on
     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
     cache and 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
\end{figure}

\mc{In applying the tools to real data Sanger data, we used the Human
  Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
  the A076941.fa {\it Arabidopsis thaliana} data set used to test WCD
  \cite{Hazelhurst08a,Hazelhurst08b}, (see Table~\ref{SangerReal}).
  We notice essentially identical results for PEACE and WCD in
  quality, both significantly better than Cap3 in Sensitivity and Type
  1 error rate, while slightly worse in the Jaccard Index and Type 2
  error.  In runtime we see some inconsistency, with PEACE showing a
  60\% runtime improvement in the first data set, but requiring 20\%
  more time in the Arabidopsis data set.  In Section~D.1 of the
  Supplementary Materials we present runtimes several more sets,
  observing that while \peace\/ appears to be significantly 
  faster on the smaller sets, \wcd\/ does overtake it for larger
  sets.}
